Benchmarking the accelerated oblique random survival forest

### 2022-06-11 paper/paper-old.Rmd
slides first draft

1. accelerated oblique random survival forests {`aorsf`}
1. original oblique random survival forests {`obliqueRSF`}
1. axis-based random survival forests {`randomForestSRC` & `ranger`}
1. axis-based conditional inference forests {`party`}
1. gradient boosted decision trees {`xgboost`}

__Note__ Presently, the original oblique random survival forest is not included in the benchmark results. This is because the original oblique random survival forest runs very slowly, turning a 10-minute benchmark into a 6 hour benchmark. I will include it once the project is more stable.

__Note__ {`xgboost`} requires careful tuning to obtain ideal performance. While we do tune the number of boosting steps in this benchmark, we do not go beyond that and instead use default setting for hyper-parameters.

@@

1. `ANOVA importance`: A method introduced by Menze et al. This is computed using an accelerated oblique random survival forest model.

1. `SHAP importance`: A method introduced by Lundberg et al. SHAP importance is widely regarded as a gold standard due to its excellent asymptotic properties. This is computed using a boosting model.

1. `Permutation importance`: A method introduced by Leo Breiman. This is computed using a standard random survival forest model.

__Evaluation of variable importance techniques__: VI techniques are compared based on their discrimination (C-statistic) between relevant and irrelevant predictor variables. The relevant predictor variables are those that have a relationship to the outcome, and the irrelevant variables are those with no relationship to the outcome. This is equivalent to assigning a binary outcome to each variable: 1 if it is relevant, 0 otherwise, and then computing a C-statistic for discrimination of the relevant versus irrelevant variables using VI as a 'prediction' for the variables.

__Details on C-statistic computation:__ The C-statistic for VI is not the same thing as the C-statistic for prediction accuracy. Since relevance of a variable is binary, the C-statistic for VI is computed using `yardstick::roc_auc`, which wraps `pROC::roc`.

@@

__Mean and standard deviation__ The multivariate normal distribution that generated values for the predictor matrix had a mean of zero and standard deviation of 1 for each variable.

__Correlation__: The correlation among $z$ variables was allowed to be non-zero as was the correlation among $x$ variables and the correlation between $x$ and $z$ variables. For all non-zero correlations, a value was selected at random uniformly between a lower and upper boundary. We set the lower boundary as -0.1 and upper boundary as +0.1. The correlations were adjusted if needed to create the nearest positive definite covariance matrix using `Matrix::nearPD()`.

__Effect sizes__: A total effect of 6 was distributed evenly across each predictor group. For example, with a total of $k$ linear effect predictors (i.e., the $x$ variables), the increase in log-hazard associated with a 1 unit increased in each $x$ variable was $6 / 25 = 0.24$, i.e., about a 27% increase in relative risk.

__Outcomes__: A censored outcome variable was generated with the `simsurv::simsurv()` function using a predictor matrix that included all relevant predictors. After the outcome was generated the interactions, non-linear transformations, and linear combinations of relevant predictors were dropped from the data.

@@

__Data generation__: There are no tests for this step, just a glimpse of the simulated set of data.

@@

__Column means__: This test verifies that the mean of each predictor is close to zero, and that roughly 45% of outcomes were censored.

@@

__Effects detected__: This test verifies that, when the oracle model is used, an effect is only detected for variables that have a relationship to the outcome.

@@

__Effect size estimation for linear effects__: This test verifies that, when the oracle model is used, the estimated effect sizes for variables with linear effects are consistent with the effect sizes used to simulate the data.

@@

__Effect size estimation for non-linear effects__: This test verifies that, when the oracle model is used, the estimated effect sizes for variables with non-linear effects are consistent with the effect sizes used to simulate the data.

@@

Overall, {`aorsf`} had the highest prognostic accuracy (__Table 1__), obtaining the highest C-statistic and scaled brier score in each of the three datasets. In addition, {`aorsf`} grew tree ensembles about 10 times as fast as {`ranger`} and generated predictions about 25 times as fast as {`party`}. Not bad! While {`aorsf`} was slower than {`randomForestSRC`}, its predictions were substantially more accurate.

__Table 1__ Prognostic accuracy and computational efficiency of machine learning algorithms applied to real and simulated data.

@@

__Table 2__: Discrimination between relevant and irrelevant predictor variables for each VI technique in . Cells that are highlighted with a light blue color indicate which variable importance technique had the best discrimination for a given sample size, correlation setting, and variable type.

@@

1. Source for `rotterdam` data: Patrick Royston and Douglas Altman, External validation of a Cox prognostic model: principles and methods. BMC Medical Research Methodology 2013, 13:33

>>>>

1.  accelerated oblique random survival forests {`aorsf`}
2.  original oblique random survival forests {`obliqueRSF`}
3.  axis-based random survival forests {`randomForestSRC` & `ranger`}
4.  axis-based conditional inference forests {`party`}
5.  gradient boosted decision trees {`xgboost`}

**Note** Presently, the original oblique random survival forest is not included in the benchmark results. This is because the original oblique random survival forest runs very slowly, turning a 10-minute benchmark into a 6 hour benchmark. I will include it once the project is more stable.

**Note** {`xgboost`} requires careful tuning to obtain ideal performance. While we do tune the number of boosting steps in this benchmark, we do not go beyond that and instead use default setting for hyper-parameters.

@@

2.  `ANOVA importance`: A method introduced by Menze et al. This is computed using an accelerated oblique random survival forest model.

3.  `SHAP importance`: A method introduced by Lundberg et al. SHAP importance is widely regarded as a gold standard due to its excellent asymptotic properties. This is computed using a boosting model.

4.  `Permutation importance`: A method introduced by Leo Breiman. This is computed using a standard random survival forest model.

**Evaluation of variable importance techniques**: VI techniques are compared based on their discrimination (C-statistic) between relevant and irrelevant predictor variables. The relevant predictor variables are those that have a relationship to the outcome, and the irrelevant variables are those with no relationship to the outcome. This is equivalent to assigning a binary outcome to each variable: 1 if it is relevant, 0 otherwise, and then computing a C-statistic for discrimination of the relevant versus irrelevant variables using VI as a 'prediction' for the variables.

**Details on C-statistic computation:** The C-statistic for VI is not the same thing as the C-statistic for prediction accuracy. Since relevance of a variable is binary, the C-statistic for VI is computed using `yardstick::roc_auc`, which wraps `pROC::roc`.

@@

**Mean and standard deviation** The multivariate normal distribution that generated values for the predictor matrix had a mean of zero and standard deviation of 1 for each variable.

**Correlation**: The correlation among $z$ variables was allowed to be non-zero as was the correlation among $x$ variables and the correlation between $x$ and $z$ variables. For all non-zero correlations, a value was selected at random uniformly between a lower and upper boundary. We set the lower boundary as -0.1 and upper boundary as +0.1. The correlations were adjusted if needed to create the nearest positive definite covariance matrix using `Matrix::nearPD()`.

**Effect sizes**: A total effect of 6 was distributed evenly across each predictor group. For example, with a total of $k$ linear effect predictors (i.e., the $x$ variables), the increase in log-hazard associated with a 1 unit increased in each $x$ variable was $6 / 25 = 0.24$, i.e., about a 27% increase in relative risk.

**Outcomes**: A censored outcome variable was generated with the `simsurv::simsurv()` function using a predictor matrix that included all relevant predictors. After the outcome was generated the interactions, non-linear transformations, and linear combinations of relevant predictors were dropped from the data.

@@

**Data generation**: There are no tests for this step, just a glimpse of the simulated set of data.

@@

**Column means**: This test verifies that the mean of each predictor is close to zero, and that roughly 45% of outcomes were censored.

@@

**Effects detected**: This test verifies that, when the oracle model is used, an effect is only detected for variables that have a relationship to the outcome.

@@

**Effect size estimation for linear effects**: This test verifies that, when the oracle model is used, the estimated effect sizes for variables with linear effects are consistent with the effect sizes used to simulate the data.

@@

**Effect size estimation for non-linear effects**: This test verifies that, when the oracle model is used, the estimated effect sizes for variables with non-linear effects are consistent with the effect sizes used to simulate the data.

@@

Overall, {`aorsf`} had the highest prognostic accuracy (**Table 1**), obtaining the highest C-statistic and scaled brier score in each of the three datasets. In addition, {`aorsf`} grew tree ensembles about 10 times as fast as {`ranger`} and generated predictions about 25 times as fast as {`party`}. Not bad! While {`aorsf`} was slower than {`randomForestSRC`}, its predictions were substantially more accurate.

**Table 1** Prognostic accuracy and computational efficiency of machine learning algorithms applied to real and simulated data.

@@

**Table 2**: Discrimination between relevant and irrelevant predictor variables for each VI technique in . Cells that are highlighted with a light blue color indicate which variable importance technique had the best discrimination for a given sample size, correlation setting, and variable type.

@@

2.  Source for `rotterdam` data: Patrick Royston and Douglas Altman, External validation of a Cox prognostic model: principles and methods. BMC Medical Research Methodology 2013, 13:33
